% Chapter X

\chapter{Biological and technical discussions} % Chapter title
\label{ch:03-01} % For referencing the chapter elsewhere, use \autoref{ch:name} 

\section{Representation of protozoan genomes}

The representation of species with sequenced genomes is traditionally skewed towards mammals and vertebrate animals \cite{hotalingGenomeSequenceEvery2021}. Many groups which are much more abundant in nature or ecologically important are underrepresented in reference genome databases \cite{davidSequencingDisparityGenomic2019}. \textit{Acanthamoeba} provides a fitting example, as they are ubiquitous in aquatic environment and form important interactions with numerous other microorganisms, yet have no chromosome quality reference available. The reference genomes of the two \textit{A. castellanii} strains generated here provide the first high quality reference in the \textit{Acanthamoeba} group and thus a valuable resource for the comparative study of amoebae.

\section{Host plasticity of intracellular bacteria}

Throughout the previous part, we have developed new approaches to detect chromatin features and quantify their changes during infection (Chap. \ref{ch:02-01}). We then applied these methods in two different infection settings: Infection of the amoeba \textit{A. castellanii} by \textit{L. pneumophila} (Chap. \ref{ch:02-02}) and of murine bone marrow-derived macrophages by \textit{S. enterica} (Chap. \ref{ch:02-03}). Although both are intracellular bacteria with similar infection strategies, they can infect completely different hosts. The size of the mouse haploid genome outclasses that of \textit{A. castellanii} by two orders of magnitude (4.3 Gbp vs 45 Mbp) and its spatial organization appears much more complex, with A/B compartmentation and intricate nested loops bridging very long distances. Despite all their differences, both unicellular and human hosts are susceptible to \textit{L. pneumophila} infection. This is most impressive knowing that human is an evolutionary dead-end for \textit{L. pneumophila} due to the absence of human-to-human transmission. The bacterial genome is therefore shaped exclusively by selective pressure in its unicellular hosts. The ability to infect multicellular hosts is probably linked to the high conservation of targeted pathways and has been attributed to the wide range of protozoan hosts infected by the bacterium \cite{molofskyDifferentiateThriveLessons2004}.

This conservation was also visible in our results, as several processes deregulated during infection are common between \textit{Legionella} and \textit{Salmonella}, such as cell cycle regulation, cytoskeleton organization, protein ubiquitination and transmembrane transport.

%----------------------------------------------------------------------------------------

\section{Combination of effects}

The analyses presented in this work focus on the description of changes happening in global chromatin structure during infection. One issue with this type of experiments is that we observe the combined effect of the pathogen activity and the host immune response. There are means to dampen one of these effects, such as the use of mutant pathogens which are unable to secrete effector proteins as control to trigger host response (as used in Chapter \ref{ch:02-03}). Although these controls do not completely emulate the pathogen activity, as it will not replicate  \cite{vogelConjugativeTransferVirulence1998} and therefore will not elicit the same immune response, they are still useful to separate the effect of the infection and pathogen exposure. In the case of \textit{L. pneumophila} (Chap. \ref{ch:02-02}), reproducing the infections with mutants for the dotA secretion system and romA methyl-transferase would allow to further isolate the chromatin changes due to the infection and romA activity.

At large scale, decoupling and deciphering the individual factors at play during infection ultimately requires the use of mutagenesis screens using Transposon insertion or CRISPR. Such approaches, only assess the effect on host survival and not chromatin changes, and to our knowledge there is no method to screen for chromatin structure modifiers. Regardless, more descriptive approaches such as the ones used in this work are still important to understand the extent of changes happening during infection. Specifically, they can still inform us on the type of structural changes that the genome undergoes and global importance of genome organization during infection.


\section{Power limitations}
% Impact of the reproducibility crisis
Results from genomic analyses are especially sensitive to the parameters and methods used. This makes reproducibility in bioinformatics of utmost importance. Much like RNA-seq, Hi-C has considerable technical variability which needs to be accounted for using multiple replicates.

It was proposed that RNA-seq experiments for differential expression analysis should comprise at least 6 replicates and ideally 12 \cite{schurchHowManyBiological2016}. While this is probably true for most omics experiments, this entails a high cost which is often the limiting factor when designing experiments in genomics.

The core issue with low replicate numbers is the lack of power to distinguish between biological variability across replicates and differences due to the condition of interest. As a consequence, when fewer replicates are used, lower effect sizes (fold changes in the case of gene expression) become undetectable. This is especially problematic when studying gene regulation, where small changes in expression could be relevant.

% Importance of software quality and standardisation
Unlike RNA-seq, where the standard for analyses is well established and most softwares can account for replicates and experimental design, most methods available for Hi-C analysis do not leverage replicate information. This restricts the power of analysis to the detection of major changes.

\section{Reproducibility and reliability challenges}

The lack of standards for Hi-C data formats and processing causes a general fragmentation of bioinformatic tools, with many redundant softwares of variable quality. One recurrent issue is the absence - or low quality - of unit tests and documentation, which are unfortunately still not regarded as standard in the computational biology community. Unit tests validate each logic block of the software using inputs with known truths, as such, they could be viewed as an equivalent to control experiments in molecular biology. Software lacking these controls is more likely to contain undetected bugs that could impact results and potentially lead to false conclusions.

Some general practices can be adopted to address these issues, such as writing comprehensive documentation, solid tests and ensure long term software maintenance, but as it stands there is little incentive to do so in academia. Such incentives could come in the form of dedicated fundings for open source bioinformatics software development, for which candidates would be evaluated on the quality of their tools rather than traditional bibliometric indicators which are poorly correlated with software quality \cite{gardnerSustainedSoftwareDevelopment2021}. Ultimately, some of these quality criteria should be also enforced globally in the publishing process to ensure that tools meet quality standards for publication, as was done for example by the Journal of Open Source Software \cite{ReviewCriteriaJOSS}.

Although adopting such practices would increase the effort and time required to develop methods, the resulting tools would be more reliable, easier to use and more widely adopted, thus benefiting the global research community in the long run. Fortunately, recent years have seen an increasing adoption of good practices in bioinformatic software. One such example is the \textit{nf-core} ecosystem backed by SciLifeLab, a public institution dedicated to open-source scientific software development \url{https://www.scilifelab.se}. The generalization of similar initiatives could mean that the quality of bioinformatics software will undergo major improvements in the forseeable future \cite{ewelsNfcoreFrameworkCommunitycurated2020,wrattenReproducibleScalableShareable2021}. The general scientific community has seen other open science successes, providing a more positive outlook on the future of scientific software quality. Notable examples include the Zenodo platform for the sharing and long term archival of scientific software and data \cite{ZenodoResearchShared}, or the non-profit NumFOCUS \cite{NumFOCUSNonprofitSupporting} which supports general scientific software development. 