% !TEX root = ../thesis-example.tex
%
\pdfbookmark[0]{Acknowledgements}{Acknowledgements}
\chapter*{Acknowledgements}
\label{sec:acknowledgements}
\vspace*{-10mm}

First, I would like to thank Romain Koszul, my thesis supervisor, to whom I am very grateful for entrusting me with severeal projects and giving me the freedom to explore my scientific interests during these three years. This has helped me evolve and allowed me to discover a variety of research questions as well as supervise students and take part in exciting collaborations. I would also like to thank members of my PhD committee and reviewers for taking the time to review the present manuscript and giving me precious advice.

Thanks to all members of the RSG lab for the exceptional atmosphere and the great times we enjoyed despite the global pandemic which changed our lives for a good part of my stay in the lab. Special mentions to Axel, Axel and Lyam for all the stimulating whiteboard discussions and long debates taking place during Chromosight's development, Pierrick and Agnès for their motivation and spreading good humor the lab, Nadège for frequently cheering me up with geeky jokes and moral support as a fellow PhD student, Charlie for entrusting me with a pipette, Théo for all these interesting late night discussions, Amaury and Jacques for being such a pleasure to work with and inspiring good computational practices in the lab.

Thanks to all our collaborators, especially Carmen and Pedro who taught me a great deal about the biology of \textit{Legionella}.

J'aimerais aussi remercier mes parents pour leur soutien et leur support tout au long de cette période.
\begin{CJK*}{UTF8}{gbsn}
感谢家人在这个困难时期欢迎我进屋。
\end{CJK*}
Finalement, merci Fleur d'avoir rendu ces trois années aussi mémorables et être parvenue à me supporter en toutes circonstances.

Ce manuscript est dédié à mon grand papa, qui a éveillé mon intérêt pour la science et a toujours stimulé ma curiosité.
