% Chapter X

\chapter{Host parasite interactions} % Chapter title

\label{ch:01-01} % For referencing the chapter elsewhere, use \autoref{ch:name} 


A large number of organisms throughout the tree of life establish stable interactions with different species. These interactions are often classified according to their perceived impact on the fitness of their members. We traditionally talk about parasitism for interactions with one-way benefits, and mutualism when the interaction has a positive impact on all parties involved. Rather than a dichotomous classification, the difference between parasitism is better viewed as a spectrum, depending on the relative benefit and cost of a relationship.

Biological interactions are observed at different scales, from nanometer-scale virophages infecting giant viruses to fungi forming mycorrhyzal networks spanning several meters \citep{Johnson1997,Selosse2006} allowing exchange of nutrients with plants root systems. These interactions shape the evolutionary trajectories of the species involved and their genomic landscapes. These changes can sometimes result in drastic transitions in the organisms' lifestyle. 

This can be the case for example with intracellular bacteria forming symbiosis with their host cells, known as endosymbionts. The \textit{Wolbachia} genus is a famous example of endosymbiotic bacteria infecting arthropod species. These bacteria are reproductive parasites which can be transmitted vertically through infection of the host female's eggs \cite{Knight2001}. Some \textit{Wolbachia} have altered the reproductive capabilities of their sexual host species to reproduce asexually by \Gls{parthenogenesis} \cite{Stouthamer1993}. This effectively removes all males from the host population, benefitting the bacterium which can only be transmitted through females. In some species infection by \textit{Wolbachia} has even become necessary for reproduction. While the bacterium takes advantage of its host reproduction, it also provides numerous advantages such as resistance to viruses in flies and mosquitoes \citep{Hedges2008,Teixeira2008} and help with vitamin synthesis in bed bugs \cite{Nikoh2014}, illustrationg the blurry line between parasitism and mutualism.

In this work, we  focus on bacterial endosymbionts. Living directly inside of their host's cytoplasm, their genomic fate is most tightly linked to their host.

\section{The evolutionary context of intracellular parasitism}

Intracellular bacteria can either be facultative or obligatory endosymbionts. Obligatory endosymbionts can only replicate inside of their host cells. This is the case of several genera of obligate intracellular bacteria, such as \textit{Rickettsia} or \textit{Chlamydia}. These parasites are unable to reproduce outside of their host and become reliant on it for most metabolic pathways. The host cytoplasm being an isolated environment, obligate intracellulars have limited opportunity for recombination with other strains. Small populations of asexual organisms unable to recombine are at the mercy of \Gls{Muller}, the progressive accumulation of mutations and loss of genetic material. They undergo a process known as genome reduction: Pathways provided by the host need no longer be encoded by the parasite and are therefore lost \cite{McCutcheon2012}. This process eventually leads to the parasite becoming completely reliant on its host for survival.

Facultative bacteria opt for a different strategy, often with larger host ranges. These bacteria can survive in the extracellular space and be transmitted between different species. The "arms race" is an analogy often used to describe the evolutionary dynamics of intracellular parasites with their hosts. Each organism evolves novel strategies to improve its own fitness at the expense of the other. This is the case for intracellular bacteria such as \textit{Legionella} or \textit{Salmonella}, which can secrete a large arsenal of effector proteins into their host's cytoplasm. Many of these proteins are redundant in the sense that they interact with the same host proteins or pathways \cite{Ghosh2017}.

Perfectly redundant genes should be subject low selective pressures, making them susceptible to genetic drift and therefore unstable \cite{Bergthorsson2007}. It is therefore thought that the functions of redundant genes in intracellular bacteria have partial overlap, such as different affinity for certain substrates or ability to function in conditions or infection stages \cite{Ghosh2017}. Selective pressure would therefore be applied on these specificities. This is likely an important phenomenon for parasites with a broad host range or variable environments.

Most intracellular bacteria incorporate genes from their hosts into their genome. Such genetic transfers are known as horizontal gene transfer (HGT) and are a major contributor to bacterial genomes, with an estimated 80\% of genes being ionvolved in HGT at some point in their history \cite{Dagan2008} More recently, HGT from bacteria to eukaryotes have also been detected in eukaryotes. Although they are much less frequent 0.04-6.49\% of genes in microbial eukaryotes \cite{VanEtten2020}), gene transfers from intracellular microorganisms to eukaryotic hosts are thought to have been involved in major shifts in environmental niche. Examples are the terrestrial colonization of plants and extremophile eukaryotes such as sea ice diatoms which acquired ice binding proteins from prokaryotes \cite{VanEtten2020}.

All these exchanges illustrate the complex evolutionary dynamics of intracellular life; genetic material can be passed not only from the host to the parasite, but also between different endosymbionts and to the host.

\section{Amoeba as a host model}

Free living amoeba are ubiquitous unicellular organisms found in soil and various bodies of water, such as rivers, lakes \cite{John1995} or even puddles \cite{Sakamoto2009}. They graze on bacterial biofilms, feeding on microorganisms by phagocytosis. This lifestyle exposes them to a large number of bacteria and viruses and they are host to many endosymbionts. 

Amoeba offer a great experimental model, as many species are easy to grow in laboratory conditions and can be used for infection experiments. Despite their extensive use as an infection model, only a few species have high quality genome assemblies available and the genomics of free living amoeba is still largely unknown. For example \textit{Acanthamoeba castellanii} genome has evidence for highly variable ploidy levels \cite{Maciver2016} and horizontally acquired genes \cite{clarke2013}. These peculiar genomic features are likely important in their interactions with endosymbionts. Indeed, high ploidy levels have been proposed as a mean for asexual amoeba to escape Muller's ratchet through homologous recombination between haplotypes \cite{Maciver2016}.

% HGT to counter muller's ratchet (Nowack2016)
Similarly, the amoeba \textit{Paulinella chromatophora} has photosynthetic organelles whose genome benefits from HGT from endosymbionts, as they counteract \Gls{Muller}. This interesting observation, provides an interesting track to explain the conservation of horizontal gene transfers in the genome of free living amoeba \textit{A. castellanii} \cite{clarke2013}.

Their long term coevolution with endosymbionts make free living amoeba an interesting model for evolutionary biology and ecology, but they are also an important organism for public health as they are the reservoir of human pathogens such as \textit{Legionella pneumophila}. Besides, many free living amoeba have a biphasic life cycle, living as trophozoite to feed and reproduce, and transforming into cysts in harsher conditions. This encystation process makes them even more important from the standpoint of public health; intracellular bacteria infecting amoeba are able to survive water chlorination or antibiotic treatments by using the encysted amoebae as shelters.

\section{\textit{Legionella pneumophila}}

\textit{L. pneumophila} is an important model for studying intracellular bacteria and infects a range of 15 species of amoebae and ciliated protozoa in the wild \cite{Rowbotham1980}. It can also infect lung macrophages of humans and other mammalians. In humans, this can cause a severe pneumonia known as \textit{Legionnaire's disease} \cite{CentersforDiseaseControlandPrevention2018}. Human to human transmission of \textit{L. pneumophila} is extremely rare \cite{Correia2016}, making infection of macrophages an evolutionary dead-end for the bacterium. \textit{L. pneumophila} is a major public health concern as it can contaminate water distribution systems and cause major outbreaks. The outbreak which lead to the identification of this bacterium and after which the bacteria was named happened at a convention of the American legion, Philadelphia, in 1976. This outbreak resulted in 182 cases, 29 of whom died. Since then, outbreaks are associated to \textit{Legionella} every year with over 32,000 cases reported between 1995 and 2005 \cite{who2020}.

When it is engulfed by a predatory cell, Legionella evades the lysosomal degradation route and survives in a "Legionella containing vacuole" (LCV). It does so using its type IV secretion system to secrete around 300 effector proteins into the host cytoplasm and rewire the host metabolic and signalling pathways. This stabilizes the pH, recruits nutrients and proteins towards the LCV and favors proliferation of \textit{L. pneumophila} in the host.

It was recently found that RomA, one of the effectors secreted by \textit{L. pneumophila} is a histone methyltransferase which can alter the histone methylation state throughout the host genome and affects the expression of a large number of genes \cite{Rolando2013}.
% Obtains nutrients by promoting host proteasomal degradation.
\section{\textit{Salmonella enterica}}

Unlike \textit{L. pneumophila}, \textit{S. enterica} only infects birds and mammals. It is also a model for intracellular bacterial infections and a major human pathogen. \textit{S. enterica} is the causative agent of typhoid fever.
